Divergent elliptical biface morphology: Distinct local retouch practice or serpent dance proxy for Elapids and Vipers?

### 2022-12-21 supp/index.qmd
<init lit>

Elliptical bifaces are prevalent at Mississippian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances. Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differ in noteworthy ways that are manifest in the associated material culture assemblages, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.

@@

The elliptical bifaces used in this study represent two of the largest samples cached at Caddo sites, and each was manufactured with a distinct raw material, providing for an empirical test of morphological differences as a function of raw material. If the elliptical bifaces were intended to represent local elapid and viper populations in the serpent dance, then their morphology may differ due to regional differences in biogeography. The sample of elliptical bifaces was also divided into two size classes; one conceptually reflective of production (large) or adult elapids and vipers, and the other with local reduction practices (small) or juvenile elapids and vipers.

Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface morphologies in the small class suggestive of relative standardization due to resharpening and/or retouch in one instance, and morphological divergence associated with adult elapids and vipers in the other. Size classes were also used to test two hypotheses related to morphological diversity; one where greater morphological variation would be apparent in the small size class due to idiosyncratic responses related to local retouch practices, and a second where greater morphological variation would be apparent in the large size class due to differences in constituent elapid and viper communities depicted by the bifaces.

@@

In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been meant to convey motion, where flake scars on one face represent muscle constriction along one edge and muscle dilation on the other, and where that pattern is inverted on the opposite face. Thus, in intentionally flipping between the two faces---a potential component of the serpent/snake dance---dancers could convey motion using a static implement. The parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.

>>>>

Elliptical bifaces are prevalent at Mississippian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances. Morphological attributes of elliptical bifaces used in Mississippian and Caddo imagery differ in noteworthy ways that are manifest in the associated material culture assemblages. Those from Caddo sites more readily articulate with the serpent dance, where the decorative motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.

@@

The elliptical bifaces used in this study come from two of the largest caches found at Caddo sites, and each was manufactured with a distinct raw material, providing for an empirical test of morphological differences as a function of raw material. If the elliptical bifaces were intended to represent local elapid and viper populations in the serpent dance, then their morphology may differ due to regional differences in biogeography. The sample of elliptical bifaces was also divided into two size classes; one conceptually reflective of production (large) or adult elapids and vipers, and the other with local reduction practices (small) or juvenile elapids and vipers.

Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface morphologies in the small size class suggestive of relative standardization due to resharpening and/or retouch in one instance, and morphological divergence associated with adult elapids and vipers in the other. Size classes were also used to test two hypotheses related to morphological diversity; one where greater morphological variation would be apparent in the small size class due to idiosyncratic responses related to local retouch practices, and a second where greater morphological variation would be apparent in the large size class due to differences in constituent elapid and viper communities depicted by the bifaces.

@@

In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been meant to convey motion, where flake scars on one face represent muscle constriction along one edge and muscle dilation on the other, and where that pattern is inverted on the opposite face. Thus, in intentionally flipping between the two faces---a potential component of the serpent/snake dance---dancers could convey motion using a static implement. The parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which can shift in similar ways to generate motion.
